### 2025-07-24 paper/preprint/appendix.tex
added text for IG

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-default}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-eval-default}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]

@@

>>>>

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-default}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

We make use of integrated gradients (IG) proposed by Sundararajan, Taly,
and Yan (\citeproc{ref-sundararajan2017ig}{2017}) to empirically
evaluate the feature protection mechanism in CT. We choose this approach
because it produces theoretically sound results, works well for
non-linear models, and remains relatively inexpensive.

IG calculates the contribution of each input feature towards a specific
prediction by approximating the integral of the model output with
respect to its input, using a set of samples that linearly interpolate
between a test instance and some baseline instance
(\citeproc{ref-sundararajan2017ig}{Sundararajan, Taly, and Yan 2017}).
This process produces a vector of real numbers, one per input feature,
which informs about the contribution of each feature to the prediction.
For example:

\begin{itemize}
\tightlist
\item
  a large positive value indicates that a feature has strong positive
  influence on the classification (i.e., increases the score for a
  class);
\item
  a small negative value indicates that a feature has weak negative
  influence on the classification (i.e., decreases the score for a
  class).
\end{itemize}

To calculate the contributions, IG compares the output to a baseline.
The selection of an appropriate baseline is an important design decision
--- it should produce a ``neutral'' prediction to avoid capturing
effects that cannot be directly attributed to the model
(\citeproc{ref-sundararajan2017ig}{Sundararajan, Taly, and Yan 2017};
\citeproc{ref-sturmfels2020visualizing}{Sturmfels, Lundberg, and Lee
2020}). To remain consistent in our evaluations, we use a baseline drawn
at random from a uniform distribution, \(\mathcal{U}(-1,1)\) all
datasets. This aligns with standard evaluation practices for IG.

We run IG on models trained on all datasets to compare their sensitivity
to features that were protected using CT:

\begin{itemize}
\tightlist
\item
  for synthetic datasets, this is always the first feature
\item
  for real-world tabular datasets, this is always \emph{age}
\item
  for MNIST, this is first five and last five rows of pixels
\end{itemize}

As IG outputs are not bounded (i.e., they are arbitrary real numbers),
it becomes a challenge to meaningfully compare IG outputs of different
models --- ones that are trained conventionally, and ones that underwent
counterfactual training. For our purposes, we observe that the direction
of contribution (sign) does not matter. If protecting a feature flips
the sign of the corresponding output, then CT „overcorrected'' for this
feature, but as long as the magnitude of its contribution decreases,
then CT achieved its objective according to our formulation. Thus, we
focus on the magnitude of IG outputs in the evaluation. Next, to
accommodate for variable ranges of outputs, we standardize the outputs.

Let \(\mathbb{g}_d\) denote the estimated IG for feature \(d\). Then in
the case of 2D synthetic datasets we find that taking the absolute value
of the outputs, \(|\mathbb{g}_d|\), and then dividing them by a
\(\max(\mathbb{g}) - \min(\mathbb{g})\) term allows us to make the most
meaningful comparison. In the case of real-world datasets we choose to
normalize the values to a 0-1 range instead.

We compare the (average) sensitivity to the features that were protected
for CT models. We again use bootstrapping (100 rounds, 2500 samples per
round) to establish the significance of our results. These are presented
in Table~\ref{tbl-ig}.

We observe substantial and statistically significant improvements for
seven out of nine datasets. For the synthetic datasets we observe
reduction in sensitivity to the protected feature in 3 out of 4 cases.
The only dataset negatively impacted by CT --- albeit with highly
variable results --- is \emph{Moon}, which clearly violates the
assumptions in Proposition 3.1. For the real-world datasets, the
reductions are over two-fold for \emph{Adult} and \emph{CH} and almost
three-fold for \emph{MNIST}. In case of \emph{Cred} we can fully prevent
the CT model from considering \emph{age} as a factor in classification.
Only for \emph{GMSC} we observe negative impacts of CT which we believe
is due to any or all of the following: a) data assumptions are violated;
b) the impact of other components of the CT objective outweighs expected
effects of feature protection; or c) the default baseline choice applied
consistently to all datasets is not appropriate for GMSC.

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]

@@

\bibitem[\citeproctext]{ref-sturmfels2020visualizing}
Sturmfels, Pascal, Scott Lundberg, and Su-In Lee. 2020. {``Visualizing
the Impact of Feature Attribution Baselines.''} \emph{Distill} 5 (1):
e22.

\bibitem[\citeproctext]{ref-sundararajan2017ig}
Sundararajan, Mukund, Ankur Taly, and Qiqi Yan. 2017. {``Axiomatic
Attribution for Deep Networks.''}
\url{https://arxiv.org/abs/1703.01365}.

### 2025-07-24 paper/sections/other/main_results.qmd
added text for IG

>>>>

We make use of integrated gradients (IG) proposed by @sundararajan2017ig to empirically evaluate the feature protection mechanism in CT. We choose this approach because it produces theoretically sound results, works well for non-linear models, and remains relatively inexpensive.

IG calculates the contribution of each input feature towards a specific prediction by approximating the integral of the model output with respect to its input, using a set of samples that linearly interpolate between a test instance and some baseline instance [@sundararajan2017ig]. This process produces a vector of real numbers, one per input feature, which informs about the contribution of each feature to the prediction. For example:

- a large positive value indicates that a feature has strong positive influence on the classification (i.e., increases the score for a class);
- a small negative value indicates that a feature has weak negative influence on the classification (i.e., decreases the score for a class).

To calculate the contributions, IG compares the output to a baseline. The selection of an appropriate baseline is an important design decision — it should produce a "neutral" prediction to avoid capturing effects that cannot be directly attributed to the model [@sundararajan2017ig; @sturmfels2020visualizing]. To remain consistent in our evaluations, we use a baseline drawn at random from a uniform distribution, $\mathcal{U}(-1,1)$ all datasets. This aligns with standard evaluation practices for IG.

We run IG on models trained on all datasets to compare their sensitivity to features that were protected using CT:

- for synthetic datasets, this is always the first feature
- for real-world tabular datasets, this is always *age*
- for MNIST, this is first five and last five rows of pixels

As IG outputs are not bounded (i.e., they are arbitrary real numbers), it becomes a challenge to meaningfully compare IG outputs of different models — ones that are trained conventionally, and ones that underwent counterfactual training. For our purposes, we observe that the direction of contribution (sign) does not matter. If protecting a feature flips the sign of the corresponding output, then CT „overcorrected” for this feature, but as long as the magnitude of its contribution decreases, then CT achieved its objective according to our formulation. Thus, we focus on the magnitude of IG outputs in the evaluation. Next, to accommodate for variable ranges of outputs, we standardize the outputs.

Let $\mathbb{g}_d$ denote the estimated IG for feature $d$. Then in the case of 2D synthetic datasets we find that taking the absolute value of the outputs, $|\mathbb{g}_d|$, and then dividing them by a $\max(\mathbb{g}) - \min(\mathbb{g})$ term allows us to make the most meaningful comparison. In the case of real-world datasets we choose to normalize the values to a 0-1 range instead.

We compare the (average) sensitivity to the features that were protected for CT models. We again use bootstrapping (100 rounds, 2500 samples per round) to establish the significance of our results. These are presented in @tbl-ig.

We observe substantial and statistically significant improvements for seven out of nine datasets. For the synthetic datasets we observe reduction in sensitivity to the protected feature in 3 out of 4 cases. The only dataset negatively impacted by CT — albeit with highly variable results — is *Moon*, which clearly violates the assumptions in Proposition 3.1. For the real-world datasets, the reductions are over two-fold for *Adult* and *CH* and almost three-fold for *MNIST*. In case of *Cred* we can fully prevent the CT model from considering *age* as a factor in classification. Only for *GMSC* we observe negative impacts of CT which we believe is due to any or all of the following: a) data assumptions are violated; b) the impact of other components of the CT objective outweighs expected effects of feature protection; or c) the default baseline choice applied consistently to all datasets is not appropriate for GMSC.
